---
title: "Effort Comparisons"
author: "Rachel Brooks"
date: "2023-05-22"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Using the CCFRP data at the drift level, this code examines the total fishes 
caught and total CPUE due to differences in total numbers of anglers fishing 
per drift

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

rm(list = ls());graphics.off()

library(tidyverse) # For data manipulation and summarization
library(plotrix) #To use standard error calculation
library(cowplot) #For grouping plots
library(scales) #For editing axis scales
library(vegan) #For multivariate analyses
library(rstatix) #For running ANCOVA
library(ggpmisc) #adding equation and r2 to figures
library(dplyr)
library(magrittr)
library(reshape2)
library(ggridges)  #length histograms

alpha=0.5
formula<-y~x
theme_set(theme_classic(base_size=16))

```

Load 'Raw' formats of the data tables. These are CSV versions of the
original CCFRP Access database tables. Column names are not changed
until tables are imported into R.

Raw drift data are used to merge fields with the Caught Fishes table -
this allows one to see all fish ever caught, even those excluded from
CPUE calculations.

Read in raw data tables:

```{r}
#1-Trip Information table
trip.data.raw = read.csv("Raw_Files/1-Trip Information.csv")

#3-Drift Information Table
drift.data.raw = read.csv("Raw_Files/3-Drift Information.csv", na = c("", "NA")) 

#4-Caught Fishes
fishes.caught.raw = read.csv("Raw_Files/4-Caught Fishes.csv")

#Fish codes to match species code with common name
fish.codes = read.csv("Raw_Files/Fish Species.csv")

```

*The next few chunks we will filter the data needed for analyses*

**Filter Trip Data:**

1.  Select columns of interest
2.  Rename columns as needed
3.  Filter out 'Areas'.
4.  Using mutate() to convert 'Area' to a factor with the full names as
    labels

```{r}

#str(trip.data.raw)

trip.data = trip.data.raw %>%
  select(c(Trip.ID, Area, Month, Day, Year.Automatic)) %>%
  dplyr::rename(Year = Year.Automatic) %>%
  mutate(Area = factor(Area, levels = c('TD','CM','TM','SP',
                                        'FN','BH','AN','PL',
                                        'BL','PB','PC','CP',
                                        'AI','LB','SW','LJ'),
                       labels = c('Trinidad','South Cape Mendocino','Ten Mile',
                                  'Stewarts Point','Farallon Islands', 
                                  'Bodega Head','Ano Nuevo', 'Point Lobos',
                                  'Piedras Blancas','Point Buchon',
                                  'Point Conception','Carrington Point',
                                  'Anacapa Island','Laguna Beach',
                                  'Swamis','South La Jolla'))) %>%
  
  droplevels()

#DEFINE CELLS TO BE EXCLUDED FROM ANALYSIS

excluded.cells = c("TDRR", "CMMM", "CMRR", "TMMM", "TMRR",
                   "FNMM","FNRR","SPMM", "SPRR", "BHMM", "BHRR",
                   "ANMM", "ANRR","PLMM","PLMN", "PLMO","PLRR",
                   "BLMM", "BLRR","PBMM", "PBRR",
                   "PCMM", "PCRR", "CPMM", "CPRR", "AIMM", "AIRR",
                   "LBMM", "LBRR", "SWMM", "SwRR", "LJMM", "LJRR")

```

**Filter Drift Data:**

(This should include drifts with no fish caught. This includes the
column 'Total.Fishes.Caught' to visually verify that 0-fish drifts are
included. This will be important for the fishing effort calculation.

1.  Get rid of drifts marked for exclusion (e.g. drift outside cell,
    overlapped with other cell, etc.)
2.  Select columns of interest
3.  Rename columns as needed
4.  Merge with trip data (this is an inner-join by default and will only
    select drifts that are contained in the filtered trip data above
    (i.e. no need to select particular 'Areas' again)
5.  Filter out the excluded cells
6.  Filter any drift that was 2 min or less in length (See CCFRP SOP
    document for this rule)

```{r}

drift.data = drift.data.raw %>%
 
  #DROP EXCLUDED DRIFTS - NO TEXT ALLOWED IN CELL
  filter(is.na(Excluded.Drift.Comment))%>% 

  #SELECT RELEVANT COLUMNS
  select(c(Drift.ID, Trip.ID, Grid.Cell.ID, ID.Cell.per.Trip,
           Site..MPA..REF., Drift.Time..hrs., Total.Fishes.Caught, 
           Total.Angler.Hrs,Total...Anglers.Fishing, Total.Fishes.Caught)) %>%   
  
  dplyr::rename(Site = Site..MPA..REF.,
                Drift.Time.Hrs = Drift.Time..hrs.,
                Total.Anglers=Total...Anglers.Fishing) %>%
  
  #MERGE WITH 'trip.data' TO GET TRIP LEVEL INFO ATTACHED TO EACH DRIFT
  merge(., trip.data, by = intersect("Trip.ID", "Trip.ID")) %>%   
  
  filter(!Grid.Cell.ID %in% excluded.cells) %>% #DROP DATA FROM EXCLUDED CELLS
  filter(Drift.Time.Hrs > (2/60)) %>% #FILTER DATA THAT IS GREATER THAN 2 MIN
  
  droplevels()

```

**Effort Calculation:**

This calculates 'Total Angler Hours Fished' per each 'ID.Cell.per.Trip'
field.

'ID.Cell.per.Trip' is our Sample unit and is the amount of fish caught
in a cell on a given day. This includes drifts that were greater than
2min and did not have fish caught.

We are not filtering total time just yet ('ID.Cell.per.Trip' time
\>2hrs). We will do that after merge with data below.

```{r}

effort.total = drift.data %>% 
  group_by(Area, Site, Year, Drift.ID, Total.Anglers, Total.Fishes.Caught) %>% 
  summarise(Total.Angler.Hours = sum(Total.Angler.Hrs)) 
#ddply needs name to be changed ('Hours' instead of 'Hrs')

```

**Fish Data:**

'fishes.caught' data frame is the full data set of fish caught
regardless of whether they were caught within CCFRP cells.

This data set will be used for gear type comparisons, but not for CPUE.

1.  Merge with raw drift data (so that no fishes are excluded)
2.  Merge with fish.codes to get Common Names
3.  Select columns of interest
4.  Rename columns for later use

```{r}

fishes.caught.all = fishes.caught.raw %>%
  merge(.,drift.data.raw,by = intersect('Drift.ID','Drift.ID'), all = F)%>%  
  # 'all = F' EXCLUDES DRIFT INFORMATION FOR DRIFTS THAT DID NOT CATCH FISH
  
  merge(.,fish.codes,by = intersect('Species.Code','Species.Code'), all = F)%>%
  #MERGE COMMON NAME; DROP UNUSED SPECIES CODES
  
  select(c(ID.Cell.per.Trip,Drift.ID,Grid.Cell.ID,Excluded.Drift.Comment,
           Total...Anglers.Fishing, Drift.Time..hrs.,Gear.Type, Station..,
           Species.Code, Common.Name, Length..cm.))%>% #CHOOSE COLUMNS OF INTEREST
  
  dplyr::rename(Exclude = Excluded.Drift.Comment,
                Drift.time.Hrs = Drift.Time..hrs.,
                Station = Station..,
                Length.cm = Length..cm.,
                Total.Anglers=Total...Anglers.Fishing)

```

**CPUE calculations:**

We need to apply all the same filters as we did with 'drift.data'

1.  Exclude fishes caught on excluded drifts
2.  Exclude fishes caught in excluded cellS
3.  Exclude fishes caught on drifts less than 2 minutes long
4.  Select columns of interest for later analysis

```{r}

fishes.caught.cpue = fishes.caught.all %>%
  
  #DROP FISHES CAUGHT IN EXCLUDED DRIFTS
  filter(is.na(Exclude)) %>% 
  
  #DROP FISHES CAUGHT IN EXCLUDED CELLS
  filter( !Grid.Cell.ID %in% excluded.cells) %>%
  
  #DROP FISHES CAUGHT ON DRIFTS LESS THAN 2 MINUTES LONG
  filter(Drift.time.Hrs > (2/60)) %>% 
  
  select(c(ID.Cell.per.Trip, Drift.ID, Grid.Cell.ID, Total.Anglers, 
           Species.Code, Common.Name, Length.cm)) %>%
  droplevels()

```

**CPUE Summary Tables:**

Next we need to aggregate the data like we would in a pivot table in
Excel

1.  Use reshape package. Much like a pivot table. This is using the
    'length' function to count fishes.
2.  Define first species column
3.  Next we merge the species count 'pivot-table' by 'IDCell.per.Trip'
    with the effort data so that we can divide catch by total adjusted
    angler hours. Here set 'all.x = T' to keep cells that we sampled but
    did not catch any fishes. These represent entire cells that were
    fished for enough time to be included, but caught no fish 3A. Use
    mutate() to create a Total CPUE variable (Sum of all species) 3B.
    Filter out ID.Cell.per.Trip's that were less than 2 Total angler
    hours (See CCFRP SOP document)
4.  Replace NA's with Zeros: These are the cells that did not catch
    fishes NOTE: Ryan F. confirmed that these are valid data points
    through 2016
5.  Divide the counts by 'Total.Angler.Hours' to get fish/angler hour or
    CPUE

```{r}

recast.counts = dcast(fishes.caught.cpue, Drift.ID + 
                Grid.Cell.ID ~ Common.Name, length, value.var = "Species.Code")

first.spp = colnames(recast.counts)[3]
  #last.spp = colnames(recast.counts)[length(recast.counts)]

Counts.per.IDcell = merge(effort.total, recast.counts, 
                          by = intersect("Drift.ID", 
                                         "Drift.ID"), all.x = T) %>% 
  
  mutate(Total = rowSums(.[which(colnames(.) == first.spp):length(.)]))
  
  #KEEP DATA WITH GREATER THAN 2 ANGLER HOURS
  #filter(Total.Angler.Hours>2) 

#DEFINE COLUMNS WITH SPECIES (E.G., COLUMNS BARRED SAND BASS:TOTAL)
spp.cols = colnames(Counts.per.IDcell)[which(colnames(Counts.per.IDcell) == 
                                      first.spp):length(Counts.per.IDcell)]

#USE THESE COLUMNS TO SELECT RELEVANT DATA AND REPLACE NA's WITH 0
Counts.per.IDcell[spp.cols] <- replace(Counts.per.IDcell[spp.cols], 
                                       is.na(Counts.per.IDcell[spp.cols]), 0)

#DIVIDE COUNTS BY EFFORT TO GET CPUE
CPUE.per.IDcell = Counts.per.IDcell
CPUE.per.IDcell[spp.cols] = (Counts.per.IDcell[, spp.cols]/Counts.per.IDcell
                                                   [, "Total.Angler.Hours"])

#EXPORT CPUE AS CSV FILE FOR SUMMARY FIGURES AND ANALYSIS
write.csv(CPUE.per.IDcell, file = 'CPUE-Per-DriftID-CCFRP-2022.csv',
                                                  row.names = FALSE)

```

Import CPUE data table (CPUE-Per-DriftID-CCFRP-2022.csv) derived from
raw data tables downloaded from CCFRP Access database. Need to transform
data into long format to calculate average CPUE for all species across
years.

```{r}
cpue<-read.csv('CPUE-Per-DriftID-CCFRP-2022.csv')

#Gather data into long format
cpue<-cpue%>%
  group_by(Area, Site, Year)%>%
  gather(species, cpue,-Drift.ID,  -Area, -Site, -Year, 
         -Total.Angler.Hours, -Total.Anglers, -Grid.Cell.ID, 
         -Total.Fishes.Caught)
#Filter species to include all fish species (total)
cpue.total<-cpue%>%
  group_by(Area, Site, Year, species)%>%
  filter(species == "Total")

#Make angler totals a factor and not numeric
cpue.total$Total.Anglers <- as.factor(cpue.total$Total.Anglers)

```

Total fishes caught per drift by number of anglers on a trip:
(1) Scatter plot - all areas data
(2) Scatter plot - separated by area 
(3) Box plot - all areas data
(4) Box plot - separated by area 

```{r, fig.height = 7, fig.width = 16}

#Raw Scatter Plots - each point is a drift

#Plot CPUE per drift by number of anglers for each sampling area
ggplot(cpue.total, aes(x = Total.Anglers, y = Total.Fishes.Caught, color=Area))+
  geom_point(size = 3, shape = 16) +
  ylab('Total fishes caught per drift')+
  xlab('Number of anglers')+
  ggtitle('Total fishes caught per drift by number of anglers on trip - all areas combined')

#Plot CPUE per drift by number of anglers for each sampling area
ggplot(cpue.total, aes(x = Total.Anglers, y = Total.Fishes.Caught, color=Year))+
  geom_point(size = 3, shape = 16)+
  ylab('Total fishes caught per drift')+
  xlab('Number of anglers')+
  facet_wrap(~Area)+
  ggtitle('Total fishes caught per drift by number of anglers on trip - separated by area')


#Box Plots

# Create boxplot using all drift-specific CPUE data
ggplot(cpue.total, aes(x = Total.Anglers, y = Total.Fishes.Caught, fill = Total.Anglers)) +
  geom_boxplot() +
  ylab('Total fishes caught per drift') +
  xlab('Number of anglers')+
  ggtitle('Total fishes caught per drift by number of anglers on trip - all areas combined')

# Create boxplot using all drift-specific CPUE data facet wrapping area
ggplot(cpue.total, aes(x = Total.Anglers, y = Total.Fishes.Caught, fill = Total.Anglers)) +
  geom_boxplot() +
  ylab('Total fishes caught per drift') +
  xlab('Number of anglers') +
  facet_wrap(~ Area)+
  ggtitle('Total fishes caught per drift by number of anglers on trip - separated by area')

```

Total fish CPUE per drift by number of anglers on a trip:
(1) Scatter plot - all areas data
(2) Scatter plot - separated by area 
(3) Box plot - all areas data
(4) Box plot - separated by area 

```{r, fig.height = 7, fig.width = 16}

#Raw Scatter Plots - each point is a drift

#Plot CPUE per drift by number of anglers for each sampling area
ggplot(cpue.total, aes(x = Total.Anglers, y = cpue, color=Area))+
  geom_point(size = 3, shape = 16) +
  ylab('CPUE')+
  labs(y=expression(CPUE~(no.~angler~hr^{-1})))+
  xlab('Number of anglers')+
  ggtitle('Total CPUE per drift by number of anglers on trip - all areas combined')

#Plot CPUE per drift by number of anglers for each sampling area
ggplot(cpue.total, aes(x = Total.Anglers, y = cpue, color=Year))+
  geom_point(size = 3, shape = 16)+
  ylab('CPUE')+
  labs(y=expression(CPUE~(no.~angler~hr^{-1})))+
  xlab('Number of anglers')+
  facet_wrap(~Area)+
  ggtitle('Total CPUE per drift by number of anglers on trip - separated by area')


#Box Plots

#Make angler totals a factor and not numeric
cpue.total$Total.Anglers <- as.factor(cpue.total$Total.Anglers)

# Create boxplot using all drift-specific CPUE data
ggplot(cpue.total, aes(x = Total.Anglers, y = cpue, fill = Total.Anglers)) +
  geom_boxplot() +
  ylab('CPUE') +
  labs(y = expression(CPUE~(no.~angler~hr^{-1}))) +
  xlab('Number of anglers')+
  ggtitle('Total CPUE per drift by number of anglers on trip - all areas combined')

# Create boxplot using all drift-specific CPUE data facet wrapping area
ggplot(cpue.total, aes(x = Total.Anglers, y = cpue, fill = Total.Anglers)) +
  geom_boxplot() +
  ylab('CPUE') +
  labs(y = expression(CPUE~(no.~angler~hr^{-1}))) +
  xlab('Number of anglers') +
  facet_wrap(~ Area)+
  ggtitle('Total CPUE per drift by number of anglers on trip - separated by area')

```


```